Learning Multiple Networks via Supervised Tensor Decomposition

### 2020-11-13 tensor_regress/NeurIPSw/camera_ready/neuripsw_tensor.tex
camera ready

% ready for submission
 \usepackage{neurips_2020}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2020}

@@

@@

\begin{abstract}
We consider the problem of tensor decomposition with multiple side information available as interactive features. Such problems are common in neuroimaging, network modeling, and spatial-temporal analysis. We develop a new family of exponential tensor decomposition models and establish the theoretical accuracy guarantees. An efficient alternating optimization algorithm is further developed. Unlike earlier methods, our proposal handles a broad range of data types, including continuous, count, and binary observations, along with available features. We apply the method to diffusion tensor imaging data from human connectome project and identify the key brain connectivity patterns associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas. Toward this end, all data and code has been made available to the public. 
\end{abstract}

@@

{\bf Notation.} We follow the tensor notation as in~\cite{kolda2009tensor}. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as $\tY\times\{\mX_1,\ldots,\mX_K\} =\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\cdots x_{j_K,i_K}^{(K)}}$,
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. The inner product between two tensors of equal size is defined as $\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}$. For ease of notation, we allow basic arithmetic operators (e.g.,\ $+, -$) and univariate functions $f\colon \mathbb{R}\to \mathbb{R}$ to be applied to tensors in an element-wise manner.

%A low-rank structure is imposed to the conditional mean of tensor observation.where unlike classical decomposition, the tensor factors $\mX_k\mM_k\in\mathbb{R}^{d_k\times r_k}$ belong to the space spanned by features $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,2,3$. The unknown matrices $\mM_k\in\mathbb{R}^{p_k\times r_k}$ (referred to as ``dimension reduction matrices'') link the conditional mean to the feature spaces, thereby allowing the identification of variations in the tensor data attributable to the side information.

 % This flexibility is important in practice. Furthermore, our low-rank model on the (transformed) conditional mean tensor effectively mitigates the curse of high dimensionality. In classical GLM, the sample size and feature dimension are well defined; however, in the tensor data analysis, we observe only one realization of an order-$K$ tensor and up to $K$ interactive feature matrices. Both the number of tensor entries and feature dimension grow exponentially in $K$. Dimension reduction is therefore crucial for prediction and interpretability. We establish the statistical convergence of our estimator, and we quantify the gain in prediction through simulations and case studies.

\section{Proposed models and motivating examples}
%For any two matrices $\mA,\mB$ with orthonormal columns of same dimension, the angle distance is defined as
%\[ \sin \Theta(\mA,\mB)=\snormSize{}{\mA^T \mB^{\perp}}=\max\left\{ \frac{\langle \mx, \my\rangle}{\vnormSize{}{\mx}\vnormSize{}{\my}}\colon \ \mx\in \textup{Span}(\mA),\ \my\in \textup{Span}(\mB^\perp)\right\}. \]
Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose the side information is available on each of the $K$ modes. Let $\mX_k=\entry{x_{ij}}\in\mathbb{R}^{d_k\times p_k}$ denote the feature matrix on the mode $k\in[K]$, where $x_{ij}$ denotes the $j$-th feature value for the $i$-th tensor entity, for $(i,j)\in[d_k]\times[p_k]$, $p_k\leq d_k$. We assume that, conditional on the features $\mX_k$, the entries of tensor $\tY$ are independent realizations from an exponential family distribution, and the conditional mean tensor admits the form
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1, \ldots, \mX_K\mM_K\}\right),\notag\\
\text{with} \ &\ \mM_k^T\mM_k = \mI_{r_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\quad \text{for all }k=1,\ldots,K.
\end{align}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns with $r_k\leq p_k$ for all $k\in[K]$, and $f(\cdot)$ is a known link function whose form depending on the data type of $\tY$. Common choices of link functions include identity link for Gaussian distribution, logistic link for Bernoulli distribution, and $\exp(\cdot)$ link for Poisson distribution.

@@

\section{Estimation algorithms}
We develop a likelihood-based procedure to estimate $\tC$ and $\mM_k$ in~\eqref{eq:decomp}. Ignoring constants that do not depend on $\Theta$, the quasi log-likelihood of~\eqref{eq:decomp} is equal to
\begin{equation}\label{eq:loglikelihood}
\small
\tL_{\tY}(\tC,\mM_1,\ldots,\mM_K)=\langle \tY, \Theta \rangle - \sum\nolimits_{i_1,\ldots,i_K} b(\theta_{i_1,\ldots,i_K})\ \text{with}\ \Theta=\tC\times\{\mM_1\mX_1,\ldots,\mM_K\mX_K\},
\end{equation}
where $b(\theta)=\theta^2/2$ for Gaussian response, $b(\theta)=\exp(\theta)$ for Poisson response, and $b(\theta)=\log(1+\exp(\theta))$ for Bernoulli response. We propose a constrained maximum quasi-likelihood estimator (M-estimator),
\begin{align} \label{eq:MLE} 
(\hat \tC, \hat \mM_1,\ldots,\hat \mM_k) &=\argmax_{(\tC,\mM_1,\ldots,\mM_K)\in \tP} \ \tL_{\tY}(\tC,\mM_1,\ldots,\mM_K),
\end{align}
where parameter space 
$
\tP=\left\{(\tC, \mM_1,\ldots,\mM_K) \ \Big| \ \mM_k^T\mM_k=\mathbf{I}_{r_k},\ \mnormSize{}{\Theta(\tC,\mM_1,\ldots,\mM_K)}\leq \alpha \right\}.
$

@@

\begin{algorithm}[!h]
\caption{Supervised Tensor Decomposition with Interactive Side Information (Simplified)}\label{alg:B}
\begin{algorithmic}[1]
\INPUT Response tensor $\tY\in \mathbb{R}^{d_1\times \cdots \times d_K}$, feature matrices $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,\ldots,K$, target Tucker rank $\mr=(r_1,\ldots,r_K)$, link function $f$, maximum norm bound $\alpha$
\OUTPUT Estimated core tensor $\hat \tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ and factor matrices $\hat \mM_k\in\mathbb{R}^{p_k\times r_k}$. 
\State Random initialization of the core tensor $\tC$ and factor matrices $\mM_k$. 
\While{Do until convergence}

%\State Obtain the factor matrix $\tilde \mM_k\in\mathbb{R}^{p_k\times r_k}$ by a GLM with link function $f$.
%\State Perform QR factorization $\tilde \mM_k=\mQ\mR$.
%\State Update $\mM_k\leftarrow \mQ$ and core tensor $\tC\leftarrow \tC\times_k \mR$.
\State Obtain $\tilde \mM_k\in\mathbb{R}^{p_k\times r_k}$ by a GLM. Orthogonalize $\tilde \mM_k$ by QR factorization, for $k \in [K]$.

\State Update the core tensor $\tC$ by solving a GLM. Rescale the core tensor $\tC$ such that $\maxnorm{\tC} \leq \alpha$. 
\EndWhile
\end{algorithmic}
\end{algorithm}

@@

\section{Numerical experiments}
We compare our supervised tensor decomposition ({\bf STD}) with three other supervised tensor methods: Higher-order low-rank regression ({\bf HOLRR}~\cite{rabusseau2016low}), Higher-order partial least square ({\bf HOPLS}~\cite{zhao2012higher}) and Subsampled tensor projected gradient ({\bf TPG}~\cite{yu2016learning}). Figure~\ref{fig:compare} shows that {\bf STD} outperforms others, especially in the low-signal, high-rank setting. As the number of informative modes (i.e., modes with available features) increases, the {\bf STD} exhibits a substantial reduction in error whereas others remain unchanged (Figure~\ref{fig:compare}b). This showcases the benefit of incorporation of multiple features. The accuracy gain in Figure~\ref{fig:compare} demonstrates the benefit of alternating algorithm -- incorporation of informative modes also improves the estimation in the non-informative modes.

>>>>

% ready for submission
% \usepackage{neurips_2020}
\usepackage[final]{neurips_2020_ml4ps}
% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2020}

@@

\author{%
  Jiaxin Hu \\
  Department of Statistics\\
  University of Wisconsin-Madison\\
  Madison, WI 53706\\
  \texttt{jhu267@wisc.edu} \\

  \And
  Chanwoo Lee \\
  Department of Statistics\\
  University of Wisconsin-Madison\\
  Madison, WI 53706\\
  \texttt{chanwoo.lee@wisc.edu} \\

  \And

  Miaoyan Wang \\
  Department of Statistics\\
  University of Wisconsin-Madison\\
  Madison, WI 53706\\
  \texttt{miaoyan.wang@wisc.edu} \\

}

@@

\begin{abstract}
We consider the problem of tensor decomposition with multiple side information available as interactive features. Such problems are common in neuroimaging, network modeling, and spatial-temporal analysis. We develop a new family of exponential tensor decomposition models and establish the theoretical accuracy guarantees. An efficient alternating optimization algorithm is further developed. Unlike earlier methods, our proposal handles a broad range of data types, including continuous, count, and binary observations, along with available features. We apply the method to diffusion tensor imaging data from human connectome project and identify the key brain connectivity patterns associated with available features. Our method will help the practitioners efficiently analyze tensor datasets in various areas. Toward this end, all data and code have been made available to the public. 
\end{abstract}

@@

{\bf Notation.} We follow the tensor notation as in~\cite{kolda2009tensor}. The multilinear multiplication of a tensor $\tY\in\mathbb{R}^{d_1\times \cdots\times d_K}$ by matrices $\mX_k=\entry{x_{i_k,j_k}^{(k)}}\in\mathbb{R}^{p_k\times d_k}$ is defined as $\tY\times\{\mX_1,\ldots,\mX_K\} =\entry{\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}x_{j_1,i_1}^{(1)}\cdots x_{j_K,i_K}^{(K)}}$,
which results in an order-$K$ $(p_1,\ldots,p_K)$-dimensional tensor. The inner product between two tensors of equal size is defined as $\langle \tY,\ \tY'\rangle =\sum_{i_1,\ldots,i_K}y_{i_1,\ldots,i_K}y'_{i_1,\ldots,i_K}$. For ease of notation, we allow basic arithmetic operators (e.g.,\ $+, -$) and univariate functions $f\colon \mathbb{R}\to \mathbb{R}$ to be applied to tensors in an element-wise manner. Besides, let $\otimes$ denote the Kronecker product of matrices.

\vspace{-0.2cm}
\section{Proposed models and motivating examples}
\vspace{-0.2cm}
%For any two matrices $\mA,\mB$ with orthonormal columns of same dimension, the angle distance is defined as
%\[ \sin \Theta(\mA,\mB)=\snormSize{}{\mA^T \mB^{\perp}}=\max\left\{ \frac{\langle \mx, \my\rangle}{\vnormSize{}{\mx}\vnormSize{}{\my}}\colon \ \mx\in \textup{Span}(\mA),\ \my\in \textup{Span}(\mB^\perp)\right\}. \]
Let $\tY=\entry{y_{i_1,\ldots,i_K}}\in\mathbb{R}^{d_1\times \cdots\times d_K}$ denote an order-$K$ data tensor. Suppose the side information is available on each of the $K$ modes. Let $\mX_k=\entry{x_{ij}}\in\mathbb{R}^{d_k\times p_k}$ denote the feature matrix on the mode $k\in[K]$, where $x_{ij}$ denotes the $j$-th feature value for the $i$-th tensor entity, for $(i,j)\in[d_k]\times[p_k]$, $p_k\leq d_k$. We assume that, conditional on the features $\mX_k$, the entries of tensor $\tY$ are independent realizations from an exponential family distribution, and the conditional mean tensor admits the form
\begin{align}\label{eq:decomp}
\mathbb{E}(\tY|\mX_1,\ldots,\mX_K) &= f\left(\tC\times\{\mX_1\mM_1, \ldots, \mX_K\mM_K\}\right),\notag\\
\text{with} \ &\ \mM_k^T\mM_k = \mI_{r_k},\ \mM_k\in\mathbb{R}^{p_k\times r_k}\quad \text{for all }k=1,\ldots,K.
\end{align}
where $\tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ is a full-rank core tensor, $\mM_k\in\mathbb{R}^{p_k\times r_k}$ are factor matrices consisting of orthonormal columns with $r_k\leq p_k$ for all $k\in[K]$, and $f(\cdot)$ is a known link function whose form depending on the data type of $\tY$. Common choices of link functions include identity link for Gaussian distribution, logistic link for Bernoulli distribution, and $\exp(\cdot)$ link for Poisson distribution.

@@

\vspace{-0.2cm}

\section{Estimation algorithms}

\vspace{-0.2cm}
We develop a likelihood-based procedure to estimate $\tC$ and $\mM_k$ in~\eqref{eq:decomp}. Ignoring constants that do not depend on $\Theta$, the quasi log-likelihood of~\eqref{eq:decomp} is equal to
\begin{equation}\label{eq:loglikelihood}
\small
\tL_{\tY}(\tC,\mM_1,\ldots,\mM_K)=\langle \tY, \Theta \rangle - \sum\nolimits_{i_1,\ldots,i_K} b(\theta_{i_1,\ldots,i_K})\ \text{with}\ \Theta=\tC\times\{\mM_1\mX_1,\ldots,\mM_K\mX_K\},
\end{equation}
where $b(\theta)=\theta^2/2$ for Gaussian response, $b(\theta)=\exp(\theta)$ for Poisson response, and $b(\theta)=\log(1+\exp(\theta))$ for Bernoulli response. We propose a constrained maximum quasi-likelihood estimator (M-estimator),
\begin{align} \label{eq:MLE} 
(\hat \tC, \hat \mM_1,\ldots,\hat \mM_k) &=\argmax_{(\tC,\mM_1,\ldots,\mM_K)\in \tP} \ \tL_{\tY}(\tC,\mM_1,\ldots,\mM_K),
\end{align}
where parameter space 
$
\tP=\left\{(\tC, \mM_1,\ldots,\mM_K) \ \Big| \ \mM_k^T\mM_k=\mathbf{I}_{r_k},\ \mnormSize{}{\Theta(\tC,\mM_1,\ldots,\mM_K)}\leq \alpha \right\}
$, and $\alpha$ is a constant. The maximum norm constraint on the linear predictor $\Theta$ is a technical condition to avoid the divergence in the non-Gaussian variance.

@@

\begin{algorithm}[htb]
\caption{Supervised Tensor Decomposition with Interactive Side Information}\label{alg:B}
\begin{algorithmic}[1]
\INPUT Response tensor $\tY\in \mathbb{R}^{d_1\times \cdots \times d_K}$, feature matrices $\mX_k\in\mathbb{R}^{d_k\times p_k}$ for $k=1,\ldots,K$, target Tucker rank $\mr=(r_1,\ldots,r_K)$, link function $f$, maximum norm bound $\alpha$
\OUTPUT Estimated core tensor $\hat \tC\in\mathbb{R}^{r_1\times \cdots \times r_K}$ and factor matrices $\hat \mM_k\in\mathbb{R}^{p_k\times r_k}$. 
\State Random initialization of the core tensor $\tC$ and factor matrices $\mM_k$. 
\While{Do until convergence}
\For { $k=1$ to $K$}
\State Obtain the factor matrix $\tilde \mM_k\in\mathbb{R}^{p_k\times r_k}$ by a GLM with link function $f$.
\State Perform QR factorization $\tilde \mM_k=\mQ\mR$. Update $\mM_k\leftarrow \mQ$ and core tensor $\tC\leftarrow \tC\times_k \mR$.
\EndFor
\State Update the core tensor $\tC$ by solving a GLM with $\textup{vec}(\tY)$ as response, $\otimes_{k=1}^K[ \mX_k\mM_k]$ as features, with link function $f$.
Rescale the core tensor $\tC$ such that $\mnormSize{}{\Theta(\tC,\mM_1,\ldots,\mM_K)}\leq \alpha$. 
\EndWhile
\end{algorithmic}
\end{algorithm}

@@

\section{Numerical experiments}

We evaluate the empirical performance of our supervised tensor decomposition (STD) through simulations. We consider order-3 tensors, where the conditional mean tensor is generated form model~\eqref{eq:decomp}.
% The core tensor entries are i.i.d.\ drawn from Uniform[-1,1], the factor matrix $\mM_k$ is uniformly sampled with respect to Haar measure from matrices with orthonormal columns. The feature matrix $\mX_k$ is either an identity matrix (i.e.,\ no feature  available) or Gaussian random matrix with i.i.d.\ entries from $N(0,1)$. The linear predictor $\Theta=\entry{\theta_{ijk}} = \tC\times\{\mM_1\mX_1,\mM_2\mX_3,\mM_3\mX_3\}$ is scaled such that $\mnormSize{}{\Theta}=1$. 
 Given the generated linear predictor $\Theta=\entry{\theta_{ijk}} = \tC\times\{\mM_1\mX_1,\mM_2\mX_3,\mM_3\mX_3\}$, the entries in the tensor $\tY=\entry{y_{ijk}}$ are drawn independently according to three probabilistic models: (a)  Gaussian model: $y_{ijk}\sim N\left( \theta_{ijk}, 1\right)$; (b)  Poisson model: $y_{ijk}\sim\textup{Poisson}\left( e^{ \theta_{ijk}}\right)$; (c) Bernoulli model:  $y_{ijk}\sim \textup{Bernoulli}\left( {e^{\theta_{ijk}}}/{1+e^{\theta_{ijk}}}\right)$.

The experiment I evaluates the accuracy when covariates are available on all modes. We set $\alpha=10, d_k=d, p_k=0.4d_k, r_k=r\in\{2,4,6\}$ and increase $d$ from 25 to 50. Our theoretical analysis suggests that $\hat \tB$ has a convergence rate $\tO(d^{-2})$ in this setting. Figure~\ref{fig:dim} plots the estimation error versus the ``effective sample size'', $d^2$, under three different distribution models. We found that the empirical MSE decreases roughly at the rate of $1/d^2$, which is consistent with our theoretical ascertainment. 
%We also observed that, tensors with higher ranks tend to yield higher estimation errors, as reflected by the upward shift of the curves as $r$ increases. Indeed, a larger $r$ implies a higher model complexity and thus greater difficulty in the estimation. 
Similar behaviors can be observed in the non-Gaussian data in Figures~2b-c.

\begin{figure}[!h]
\centering
\includegraphics[width=14cm]{dim_alternative_w.pdf}
\vspace{-.5cm}
\caption{(a)-(c): Estimation error against effective sample size. The dashed curves correspond to $\tO({1/d^2})$. (d)-(f): Performance comparison under stochastic block models. The $x$-axis represents the number of blocks in the networks. The response tensors are generated from Gaussian (a, d), Poisson (b, e) and  Bernoulli  (d, f) models . }\label{fig:dim}
\end{figure}

The experiment II investigates the capability of our model in handling correlation among coefficients. We mimic the scenario of brain imaging analysis. A sample of $d_3=50$ networks are simulated, where each network measures the connections between $d_1=d_2=20$ brain nodes. We simulate $p=5$ features for the each of the 50 individuals. These features may represent, for example, age, gender, cognitive score, etc. Recent study has suggested that brain connectivity networks often exhibit community structure represented as a collection of subnetworks, and each subnetwork is comprised of a set of spatially distributed brain nodes. To accommodate this structure, we utilize the stochastic block model~\citep{abbe2017community} to generate the effect size. Specifically, we partition the nodes into $r$ blocks by assigning each node to a block with uniform probability. Edges within a same block are assumed to share the same feature effects, where the effects are drawn i.i.d.\ from $N(0,1)$.

Figure~\ref{fig:dim}(d)-(f) compares the MSE of our method with a multiple-response GLM approach. The multiple-response GLM is to regress the dyadic edges, one at a time, on the features, and this model is repeatedly fitted for each edge. As we find in Figure~\ref{fig:dim}(d)-(f), our tensor regression method achieves significant error reduction in all three data types considered. The outperformance is substantial in the presence of large communities; even in the less structured case ($\sim 20/15=1.33$ nodes per block), our method still outer-performs GLM. The possible reason is that the multiple-response GLM approach does not account for the correlation among the edges, and suffers from overfitting. In contrast, the low-rankness in our modeling incorporates the shared information across entries.

The experiment III compares {\bf STD} with three other supervised tensor methods: Higher-order low-rank regression ({\bf HOLRR}~\cite{rabusseau2016low}), Higher-order partial least square ({\bf HOPLS}~\cite{zhao2012higher}) and Subsampled tensor projected gradient ({\bf TPG}~\cite{yu2016learning}). Figure~\ref{fig:compare} shows that {\bf STD} outperforms others, especially in the low-signal, high-rank setting. As the number of informative modes (i.e., modes with available features) increases, the {\bf STD} exhibits a substantial reduction in error whereas others remain unchanged (Figure~\ref{fig:compare}b). This showcases the benefit of incorporation of multiple features. The accuracy gain in Figure~\ref{fig:compare} demonstrates the benefit of alternating algorithm -- incorporation of informative modes also improves the estimation in the non-informative modes.
